---
title: "Install R and RStudio"
output: 
    html_document
---

```{r include=FALSE}

library(png)
library(grid)

```

We need to install two things onto your computer. The first is **R**, the statistical software package we will use (R is also the name of the programming language we use in the software). After that, we need to install **RStudio**, which is a front end program that lets you write R code, view plots and do many other useful things. 

If you have a previous installation of **R** and **RStudio** that you have not used much, please remove these first from your computer. Then download the current versions as per the instructions below.

<br>

## Install **R** 

Download the **R** installer from <a href="https://cran.r-project.org/" target="_blank">https://cran.r-project.org/</a> by clicking on the link for your operating system. 

Go to the Windows or Mac section in the instructions in this document below, as appropriate.

```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/Rinst.png")
grid.raster(img)

```


<br>

### 1. Windows

1a. Click *install R for the first time*.

```{r echo=FALSE, fig.height=4, fig.width=8}

img <- readPNG("pics/Rinst2.png")
grid.raster(img)

```

1b. Use the download link at the top and save the file. 

*Make sure the installer is for R version 4.0.5.*


```{r echo=FALSE, fig.height=4, fig.width=8}
img <- readPNG("pics/Rinst3.png")
grid.raster(img)
```

1c. Run the installer (double click). Default settings are fine. If you do not have admin rights on your laptop, then **ask IT**. If this is the case, it is important to ask them to give you full permissions to the **R** directories. Without this, you will not be able to install **R** packages. 



### 2. Mac

2a. Check the version of your Mac OS. 

2b. If you have Mac OS 10.13 (High Sierra), 10.14 (Mojave), 10.15 (Catalina) or 11 (Big Sur), click the download link for **R** version 4.0.5.

If you have Mac OS version 10.11 (El Capitan) or 10.12 (Sierra), click the download link for **R** version 3.6.3.nn.


```{r echo=FALSE, fig.height=6, fig.width=8}

img <- readPNG("pics/Rinst4.png")
grid.raster(img)

```


2c. Run the installer (double click). Default settings are fine. If you do not have admin rights on your laptop, then **ask IT**. If this is the case, it is important to ask them to give you full permissions to the **R** directories. Without this, you will not be able to install **R** packages. 


<br>

## Install RStudio

1. Wait until the **R** installer has finished.

2. Download **RStudio** <a href="https://rstudio.com/products/rstudio/download/#download" target="_blank">https://rstudio.com/products/rstudio/download/#download</a>.


The page should recommend the right file based on your system. If not, scroll down and download the installer for the appropriate free, **RStudio** Desktop version.

```{r echo=FALSE, fig.height=6, fig.width=8}

img <- readPNG("pics/InstallRstudio.png")
grid.raster(img)

```

<br>

## Check R and RStudio are working

1. Open **RStudio**. It should look like the image below.

2. We are going to try to run the following code. Follow the instructions below. 

```{r eval=FALSE}

install.packages("ggplot2")
library(ggplot2)
p <- ggplot(mtcars, aes(wt, mpg))
p + geom_point(aes(size = qsec, colour = factor(cyl)))

```

 a. Copy the code above and paste it in the Console window in **RStudio**, by the `>` sign. Before you hit Enter, it will look like this.

```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/RunTestPre.png")
grid.raster(img)

```

 b. Hit Enter. The code will run (it might take a minute or so) and there will be a plot created in the plot window at the bottom right.

```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/RunTestPost.png")
grid.raster(img)

```

Your console may look a bit different to mine but, if you have a plot (bottom right) similar to mine, all is well, you have installed **R** and **RStudio** successfully. 

c. **Please take a screenshot and email it to me.**

<br>

## Zoom configuration (Mac users)

If you are a *Mac* user, there are two Zoom settings that need to be checked.

1. Remote Control

1a. Open Zoom. In the top right corner, click your personal icon and go to *My Profile*. This will take you to your Zoom profile in a web browser.

1b. On the left, go to *Settings*. Then go to *In Meeting (Basic)*.


```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/ZoomMac1.png")
grid.raster(img)

```

1c. Scroll down and find *Remote Control*. Please make sure both settings are turned on, as shown below, and then log out.


```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/ZoomMac2.png")
grid.raster(img)

```


2. Permissions for Zoom web client

2a. On your Mac, go to *System Preferences* (Apple Menu > System Preferences ...).

2b. Click *Security & Privacy* and go to the *Privacy* tab.

2c. Scroll down on the left to *Screen Recording*.


```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/ZoomMac3.png")
grid.raster(img)

```


2d. Click the lock in the bottom left corner and enter your Mac password to open the lock, so you can make changes.

2e. Tick the box next to Google Chrome.

2f. Click the lock to close it. You should not need to make any more changes.

(The info on these settings is here: https://support.zoom.us/hc/en-us/articles/360016688031)


<br>

## Set up your screen (optional)

Setting up your screen well from the start will help get you coding fast. We strongly recommend you try the setups recommended below. The *lessons* look a lot like **RStudio**, and it is easy to get confused between your own code and the lesson code. To differentiate the two visually, we recommend you change your **RStudio** theme to something with a dark background. 

1. In **RStudio** go to Tools > Global Options.
2. Choose 'Appearance'.
3. Under 'Editor theme' chose 'Idle Fingers' (or another dark theme).

If you have dual monitors, you're all set. Put **RStudio** on one and everything else on the other. If you have a single monitor, then you need to be a little more organised.  

1. Put **RStudio** on the left half of your monitor.
2. Inside **RStudio**, grab the vertical bar between the Console and the Files/Plot windows and squish it to the right, until you mostly just see the Console. 
3. Put Zoom in the top left corner, taking up about a third of the vertical space. Sometimes I will ask you to go full screen on Zoom. Practise going from full screen to top right corner.
4. Put your browser in the bottom right, taking up the other two-thirds of the vertical space.

```{r echo=FALSE, fig.height=5, fig.width=8}

img <- readPNG("pics/ScreenSetup.png")
grid.raster(img)

```


<br>

**Author**: Gordana Popovic

**Last updated**: `r format(Sys.time(), '%d %B, %Y')`
