---
title: "Plotting in R with ggplot2"
output: html_document
fig_width: 3 
fig_height: 2 
---

# Import data to plot
We are going to plot some data about cars that is contained in the [`cars93.csv`](https://www.dropbox.com/s/5cix5ijqrc8lv5a/cars93.csv?dl=0) file. First we import the data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=TRUE, include=FALSE}
cars <- read.csv("../data/cars93.csv")
```

```{r eval=FALSE, include=TRUE}
cars <- read.csv("cars93.csv")
```

Let's have a quick look at the data first.
```{r}
head(cars)
```

```{r}
summary(cars)
```


# Install and load ggplot2

The `ggplot2` package is in the `tidyverse`, which we have already installed. If you need to install it on another computer without installing the whole tidyverse, you will know form last lesson how to do this.

```{r eval=FALSE}
install.packages("ggplot2")
```

Remember, when you are using ggplot2 (or any package) in any R session, you have to tell R to load the ggplot2 library before you can access its commands. You have to do this every time you open a new session of R.

```{r load2, message=FALSE, warning=FALSE}
library(ggplot2)
```

 
# Plot types

## Two continuous variables (scatterplot)

The basic elements of a command in ggplot2 are data `cars`, aesthetics `aes(x,y)`, and plot type `geom_point()`. 

```{r scatterplot1, fig.height = 3, fig.width = 5}
ggplot(cars, aes(x = weight, y = cityMPG)) + 
  geom_point()
```
 
Can you see your plot? Did you get an error `Error in plot.new() : figure margins too large`? These both happen because the plots come up in a plane on the left that you may have squished is you have only one screen. Time to un-squish the left plane. 
 
Wanna be fancy? Add a trend line with `+geom_smooth()`.

```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(weight, cityMPG)) + 
  geom_point() + 
  geom_smooth()
```

I have not used `x=weight` and `y=cityMPG` this time, if I don't specify, then `ggplot` will assume the first variable is the `x` variable, and the second is `y`,

What if we want different colours by car type? All you go is add a `color =` statement inside `aes()`.

```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(weight, cityMPG, color = type)) + 
  geom_point()
```

We can add smooth lines with `geom_smooth()`, but it's a bit messy here, so I will add straight lines with `method="lm"`. Also, I don't want shading, so I will add `se=FALSE`.


```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(weight, cityMPG,color=type)) + 
  geom_point() + 
  geom_smooth(se=FALSE,method="lm")
```


## One continuous variable (histogram or boxplot)

Histograms are easy, we just need one variable inside `aes()`, if we want just one histogram.

```{r boxplot1, fig.height = 3, fig.width = 5}
ggplot(cars, aes(x=cityMPG)) + 
  geom_histogram()
```

Change to boxplot by changing the plot type and specifying `y` variable. 

```{r hist1, fig.height = 3, fig.width = 5}
ggplot(cars, aes(y=cityMPG)) + 
  geom_boxplot()
```

## One continuous variable and one categorical variable (boxplot)


We can do box plots by type by setting type as the `x` variable. 
```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(x=type,y=cityMPG)) + 
  geom_boxplot()

```


And add colours by specifying either color (as before) or fill.
```{r boxplot2, fig.height = 3, fig.width = 5}
ggplot(cars, aes(type,cityMPG,fill=type)) + 
  geom_boxplot()

```

## One categorical variable (barplot)

To plot one categorical variable `type` we can make a barplot with `geom_bar()`.
```{r}
ggplot(cars, aes(type)) +
    geom_bar()
```

You can make the categories different colours with the `fill=` option using the same variable again.
```{r}
ggplot(cars, aes(type,fill=type)) +
    geom_bar()
```

## Two categorical variables (barplot)

We have two options for bar plots for two categorical variable. The defaults is a stacked bar plot. You just add the second variable as the `fill=` option.
```{r}
ggplot(cars, aes(type,fill=fuelEco)) +
    geom_bar()
```

If you prefer a side by side bar chart, you just add the option `position = "dodge"`.
```{r}
ggplot(cars, aes(type,fill=fuelEco)) +
    geom_bar(position = "dodge")
```

# Customisation

## Labels

You can change the default x and y labels with `xlab()` and `ylab()`, and add a title with `ggtitle()`. This is demonstrated on the scatterplot from earlier.
```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(weight, cityMPG,color=type)) + 
  geom_point() +
  ylab("Miles per gallon") +
  xlab("Weight (kg)") +
  ggtitle("Mileage of cars v.s. their weight")
```

## Themes 

Themes let you easily change the look of your graph with one simple function. There are lots of themes available, the most commonly used are `theme_bw()` and `theme_classic()`.

```{r , fig.height = 3, fig.width = 5}
ggplot(cars, aes(weight, cityMPG,color=type)) + 
  geom_point() +
  ylab("Miles per gallon") +
  xlab("Weight (kg)") +
  ggtitle("Mileage of cars v.s. their weight") +
  theme_classic()
```

# Challenges

We will use the `CO2` data, which is stored in R and can be accessed by typing `CO2` (into your console), no need to import. Make sure all of your plots have appropriate titles and axis labels. Use `ggplot2` for this challenge.

1. How many variables are in the CO2 data set and what kinds are they?

Hint:
```{r include=TRUE, eval=FALSE}
str(CO2)
```

2. Create a histogram of the variable `uptake`.
3. Create a boxplot of `uptake` by `Treatment`, with different colours for each treatment.
4. Create a scatter plot of `uptake` by `conc`, with different colours and symbols by Treatment.
5. Add a smooth line to the previous plot with `geom_smooth()` to better see the patterns.

Make sure the script file you have explains what each line of code does with comments and save it in your project directory. 

# Resources

We have covered the basic functionality of `ggplot2`, but there are lots of bells and whistles available. A good place to find out more about is the [cheatsheet](https://rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf). Also, have a look at the extensive [R graph gallery](https://www.r-graph-gallery.com/) for code for every possible plot you can imagine.

**Authors**: Gordana Popovic & Mark Donoghoe  

**Last updated**: `r format(Sys.time(), '%d %B, %Y')`