---
title: "Importing data, Packages and the tidyverse"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing data

Before we can run any data analyses or create figures in R, we need to import that data into R. 

## How to import an existing dataset
1. Close your R session, we want to make sure you can do this from scratch.
2. Download the [`Sessile.csv`](https://www.dropbox.com/sh/5biuat3iusltvwe/AAAQ0HTe3OlKl3DiPbvdlRJma?dl=0) dataset and put it in your project directory (remember we created a project for this course earlier and you chose a directory). 
3. Open `Sessile.csv` and have a look: it has 60 rows and 6 columns. The Richness, Temp, Depth and Abund columns are numeric, and the others are categorical. 
4. Open your project in RStudio by double clicking the .RProj file (a little blue cube with an R in it).
5. Open a script file to use (and save it).
6. You can now use the `read.csv()` command to read the data into R.

```{r include=FALSE}
sessile <- read.csv("../data/Sessile.csv")
```

```{r eval=FALSE}
sessile <- read.csv("Sessile.csv")
```

This code reads the contents of `Sessile.csv` and saves it as a data frame called `sessile`. Notice the file name has a `.csv` and is in double quotes. 

## Checking and manipulating imported data

We can then have a look at what RStudio has imported using the `str()` function.

```{r}
str(sessile)
```

The variables seem to have been imported in the right format. The summary function will give you a little more detail about the variables in your data. 

```{r}
summary(sessile)
```


We can see that Richness varies between 38 and 81. As Copper has been read into R as a character variable, `summary` only tells us that there are 60 values in the variable.  We know (because it is our data!) that it has three categories. If a variable is a categorical variable with a particular set of values, it's usually a good idea to make it a *factor* variable. 

```{r}

sessile$Copper <- factor(sessile$Copper)

str(sessile$Copper)

```


With the variables in the right format, you can then use this data in R Studio. For example, to calculate the mean and standard deviation of Richness:

```{r}
mean(sessile$Richness)
sd(sessile$Richness)
```

To look at the first three rows of the data:

```{r}
sessile[1:3,]
```

A shortcut way to do this is using the `head` function, which gives you the first 6 rows by default.

```{r}
head(sessile)
```

## Creating data to import

While you may know how to create a vector and a small data frame in R, it's really not convenient to enter data directly into R when you are creating a larger dataset to work with later.  Instead, it's much easier to work in a spreadsheet application, such as Excel or Numbers.  Data entry in a spreadsheet is simple and it's very easy to save the completed data set in a format readable by various programs.  

### What should a dataset look like?
Basically, a dataset should have separate variables in each column. Each row should contain the values in each variable for an individual case. So, for example, in a small study on educational test scores in a school, the dataset may look like: 

| studentID | gender | mathscore | readingscore |
|-----------|--------|-----------|--------------|
| 1         | male   | 78        | 87           |
| 2         | female | 92        | 99           |
| 3         | male   | 87        | 93           |
| 4         | male   | 70        | 73           |
| 5         | female | 94        | 95           |
| 6         | female | 90        | 83           |
| 7         | female | 85        | 72           |
| 8         | male   | 90        | 93           |


So, student 1 is a male who achieved 78 on the maths test and 87 on the reading test.

### General tips for data entry
In general, in a dataset, each column contains one variable and each row contains one case.
Keep your data on one worksheet within the spreadsheet. Unless you are working with an extremely complex dataset (in which case, you wouldn't be doing data management with a spreadsheet program anyway), it's much easier for you to keep everything on the same worksheet, rather than work with separate ones.

### Variable Names

- Keep the names simple and descriptive.  
- Use only lower and upper case English letters, numbers and punctuation marks.  If you begin your variable name with a number or a punctuation mark, R will prefix the the variable name with an "X" when you import the .csv file.
- In R, a variable named mathscore is not the same as a variable named mathScore. That is, in the same dataset, you could have one variable named mathscore and a different one named mathScore (although this might be a bit confusing!).
- If you need to document what the variables are, keep the descriptions in a different file.
- For string (character) variables, spelling and capitalisation matters. Entering "male" and "Male" will result in two different categories.
- In your spreadsheet program, save the file in .csv format (comma separated format).  This is a general format that is readable by many statistical programs.  
- More information is available [here](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).  

### Saving the file
All the files for this course will be in the .csv format. This is a general format that is readable by many programs. It allows only one sheet or data to be saved. To save you data as a .csv in Excel go to File -> Save As, chose the directory you want to save the file to (your Project directory), chose a file name, then under the file name will be a box for file type, chose CSV (Comma delimited) (*.csv).

# Challenges
1. Create a data set in Excel with at least three columns and ten rows. Have at least one numeric and one categorical column. 
2. Save the data as a .csv file in your project directory.
3. Read your data set into R and check that everything is okay.

Make sure you use comments to explain what each line of code in your script file does. Save the file in your project directory. 



# Installing packages and the tidyverse package

Once upon a time there was only base R. Code was ugly and impossible to read, plots were ugly and difficult to make. Then one day, the tidyverse was created, and R became usable by mere humans. Because you are learning R in a post tidyverse time, you will likely not know any base R (except the tiny bit I taught you already). From here on we will all live happily ever after in the tidyverse.

Packages in R are just bits of extra code other people have written that you can use. You will be installing packages all the time in your R travels, they are what makes R so useful and flexible. It is very easy to do. You need to know the name of the package (in this case tidyverse) and then you can just call the `install.packages()` function. You only have to do this once on each computer.

```{r eval=FALSE}
install.packages("tidyverse")
```

This will take a little while, the tidyverse is vast. The main packages we will be using that are inside the tidyverse are `tidyr`, `dplyr` for manipulating data, and `ggplot2` for plotting.

To be able to use a package we have installed we need to load the package using the `library()` command. Loading a package must be done every time you open R/RStudio.
```{r}
library(tidyr)
```

## The Pipe

One of the great inventions of the tidyverse is the Pipe ( `%>%` ). It lets us organise code in a way that is more readable. You can find it in the `tidyr` package we loaded above.

Say I want to take the numbers 1 to 5, *then* take the sine, *then* take the absolute value, *then* the square root. In base R the code for that looks like this. 

```{r}
x=1:5
sqrt(abs(sin(x)))
```

This is okay, but a bit hard to read, and it's a bit unclear which bracket belongs where. Using the pipe (and reading the pipe as "then") we can write the same operation more naturally.

```{r}
x %>% 
  sin %>% 
  abs %>% 
  sqrt
```

This will make our life easier when our code gets more complicated. 

## Challenges


1. Install and load the `lubridate` R package. We won't use the package in this course, it's just for practice. It is useful for working with dates and times.


## Challenge solutions

### Importing data

This will differ for everyone, so no solutions.

### Packages

1. Install and load the `lubridate` R package.

```{r eval=FALSE}
install.packages("lubridate")
library(lubridate)
```


**Author**: Gordana Popovic

**Last updated**: `r format(Sys.time(), '%d %B, %Y')`

