---
title: "Packages and the tidyverse"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Installing packages and the tidyverse package

Once upon a time there was only base R. Code was ugly and impossible to read, plots were ugly and difficult to make. Then one day, the tidyverse was created, and R became usable by mere humans. Because you are learning R in a post tidyverse time, you will likely not know any base R (except the tiny bit I taught you already). From here on we will all live happily ever after in the tidyverse.

Packages in R are just bits of extra code other people have writtin that you can use. You will be installing packages all the time in your R travels, they are what makes R so useful and flexible. It is very easy to do. You need to know the name of the package (in this case tidyverse) and then you can just call the `install.packages()` function. You only have to do this once on each computer.

```{r eval=FALSE}
install.packages("tidyverse")
```

This will take a little while, the tidyverse is vast. The main packages we wiil be using that are inside the tidyverse are `tidyr`, `dplyr` for manipulating data, and `ggplot2` for plotting.

To be able to use a package we have installed we need to load the pakage using the `library()` command. Loading a package must be done every time you open R/RStudio.
```{r}
library(tidyr)
```

## The Pipe

One of the great inventions of the tidyverse is the Pipe ( `%>%` ). It lets us organise code in a way that is more readable. You can find it in the `tidyr` package we loaded above.

Say I want to take the numbers 1 to 5, *then* take the sine, *then* take the absolute value, *then* the square root. In base R the code for that looks like this. 

```{r}
x=1:5
sqrt(abs(sin(x)))
```

This is okay, but a bit hard to read, and it's a bit unclear which bracket belongs where. Using the pipe (and reading the pipe as "then") we can write the same operation more naturally.

```{r}
x %>% sin %>% abs %>% sqrt
```

This will make our life easier when our code gets more complicated. 

## Challenges

```{r  include=FALSE}
x=(1:6)^6
```

1. Install and load the `DHARMa` R package.

**Author**: Gordana Popovic

**Last updated**: `r format(Sys.time(), '%d %B, %Y')`

